---
title: "Survival Analysis: Kaplan Meier estimation"
author: "Felix Eßer"
date: "8/26/2020"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

```{r load packages, message=FALSE}
library(ggfortify) # to plot survival curves with ggplot2
library(survival)
library(survminer) # different package to plot survival curves with ggplot
library(tidyverse)
```


## Basic Kaplan-Meier-Estimator

$$ \hat{S}(t) = \prod_{t_i \leq t}(1 - \hat{q_i}) = \prod_{t_i \leq t}\left( 1 - \frac{d_i}{n_i}\right)$$
where $n_i$ is the number of subjects at risk at time $t_i$, and $d_i$ is the number of individuals who fail this time. 

# Inspect data
```{r dataset}
data(veteran)
str(veteran)
```

The veteran dataset contains data from a randomized trial of two treatment regimens for lung cancer. In the documentation of the veteran dataset the following information can be found:

trt:	1=standard 2=test
celltype:	1=squamous, 2=smallcell, 3=adeno, 4=large
time:	survival time
status:	censoring status
karno:	Karnofsky performance score (100=good)
diagtime:	months from diagnosis to randomisation
age:	in years
prior:	prior therapy 0=no, 10=yes


At first, a survival object has to be created with the `Surv()` function. This object contains the time to event for each person with additional information of censoring (indicated with a '+'-sign.
```{r create survival object}
surv_obj <- with(veteran, Surv(time = time, event = status)) # create survival object
head(surv_obj, 20)
```

The next step is to use this object inside the `survfit()` function and produce the Kaplan-Meier estimate. Printing the fitted model results in overall statistics of the model.
```{r kaplan-meier estimate}
km_model <- survfit(Surv(time = time, event = status) ~ 1, data = veteran)
km_model
```

The median survival time is 80 days and the 95% confidence interval ranges from 52 to 105 days. 

Using the `summary()` function, for every time point the model's data can be printed. By aid of the `times` argument, it is possible to control which time points are printed.

```{r km summary}
summary(km_model, times = c(1:10, 50*(1:5)))
```



### Different confidence intervals
It is also possible to compute confidence intervals based on the complementary log-log transformation. This has to be explicitly specified.
```{r log-log CI}
km_log <- survfit(Surv(time = time, event = status) ~ 1,
                  conf.type = "log-log",
                  data = veteran)
km_log
```


As can be seen in comparison to the output from the `km_model` object, the 95% confidence intervals are a bit different. 



### Plots
To obtain a quick and dirty plot of the model, the `R base` plotting system can be used, which can also be modified.
```{r base r km-plot}
plot(km_model, xlab = "Time", main = "Plot of Kaplan Meier estimate")
```

Is the aim to produce a more professional plot, the package `ggplot2` can be used. Here, there are two possibilities: First, the object `km_model` is an object of the class *survfit* which can be recognized by the function `autoplot()` of the `ggplot2` package. Therefore, executing the call `autoplot(km_model)` would result in a decent plot which can be further modified. In order that this works, the package `ggfortify` has to be loaded. Second, a plot from ggplot can be produced *by hand*. Unfortunately, it is not possible to plot the confidence interval as a stepped ribbon. 
Next, the first procedure is demonstrated.

```{r ggplot km-plot}
km_table <- summary(km_model)$table # save median
autoplot(km_model,
         conf.int.alpha = 0.2, # opacity of confidence interval
         censor.colour = "blue", # change colour of censor sign
         censor.size = 5,
         xlab = "Time", ylab = "Survival probability", main = "Kaplain-Meier estimate") +
  geom_hline(yintercept = 0.5, linetype = 3) +
  geom_segment(aes(x = km_table["median"], y = 0,
                   xend = km_table["median"], yend = 0.5),
               colour = "green", linetype = "dashed") +
  geom_segment(aes(x = km_table["0.95LCL"], y = 0,
                   xend = km_table["0.95LCL"], yend = 0.5),
               colour = "blue", linetype = "dashed") +
  geom_segment(aes(x = km_table["0.95UCL"], y = 0,
                   xend = km_table["0.95UCL"], yend = 0.5),
               colour = "blue", linetype = "dashed") +
  labs(caption = "The green line indicates median survival time for 50% at 80 days. The blue lines represent the lower and upper 95% \nconfidence interval limits at 52 and 105 for the median, respectively.") +
  theme(plot.caption = element_text(hjust = 0)) #+
  #theme_light(plot.caption = element_text(hjust = 0))
```


With the package `survminer`it is possible to plot different kinds of survival curves with the `ggplot` plotting system.
```{r ggplot survminer probability}
ggsurvplot(km_model)
```




Using the argument `fun` it is possible to plot the survival probability in percentage (`"fun = "pct"`), cumulative events (`"fun = "event"`) or the cumulative hazard function (`fun = "cumhaz"`).

```{r ggplot survminer}
standard_plot <- ggsurvplot(km_model)
pct_plot <- ggsurvplot(km_model, fun = "pct")
event_plot <- ggsurvplot(km_model, fun = "event")
cumhaz_plot <- ggsurvplot(km_model, fun = "cumhaz")
arrange_ggsurvplots(list(standard_plot, pct_plot, event_plot, cumhaz_plot),
                    nrow = 2, ncol = 2,
                    title = "Different survival curves produces with survminer")
```



### Different Estimator

It is also possible to estimate the survival function by the means of a different estimator. This estimator is often called *Nelson-Altschuler* estimator or *Fleming-Harrington* estimator. 
In `R` this estimator can be used, specifying the `type = "fh"` argument in the `survfit()` function

```{r fh-estimator}
fh_model <- survfit(Surv(time = time, event = status) ~ 1, type = "fh", data = veteran)
summary(fh_model, times = c(1:10, 50*(1:5)))
```


### Median Follow-Up time
The median follow-up time is one measure of the quality of a study. It captures how long, on average, participants have been followed. The definition of the median follow-up time is ambiguous. One approach is known as the 'reverse' Kaplan-Meier in which the Kaplan-Meier estimate is estimated using the reversed censoring indicators. This means that the event is for example not death but the absent of death. A censored case is a case with an event. Through this procedure the 'potential' median survival time can be calculated. This is only reasonable if a certain amount of people have censored values. Otherwise, the median of the time variable has to be used. 

```{r reverse km}
# here no interpretable results are obtained
veteran$status_followup <- 1 - veteran$status
rev_km <- survfit(Surv(time = time, event = status_followup) ~ 1, data = veteran)
```


## Left-Truncation
Has to be added later...



## Comparison of Survival Functions

Null-hypothesis: $H_0: S_1(t) = S_0(t)$. Ther alternative hypothesis can be formulated two-sided or one-sided. Sadly it is not that simple and the hypotheses have to be reformulated because survival distribution can differ in various forms (crossing, difference starts only at a specific time point and so on)

Here, two possibilites are discussed to compare two (or more) survival functions. These tests belong to the G-rho family or Harrington-Fleming-tests. 
In the following, the two treatment groups of the veteran dataset are compared in regard to their survival functions.

```{r grouped km-plot}
autoplot(survfit(Surv(time = time, event = status) ~ trt, data = veteran)) +
  scale_fill_manual(name = "Experimental condition", labels = c("Standard", "Test Treatment"),
                    values = c(2, 4)) + # defines colors
  scale_color_manual(name = "Experimental condition", labels = c("Standard", "Test Treatment"),
                     values = c(2, 4)) +
  theme_light()
```




The first test is the log-rank or Mantel-Haenszel test. 

```{r log-rank test}
survdiff(Surv(time = time, event = status) ~ trt, # comparing both treatment groups
         data = veteran,
         rho = 0) # rho = 0 is the default value; scalar parameter
```

The log-rank test yielded a non-significant result, $\chi^2(1) = 0, p = .9$.

Through the argument `rho = 1` a modified test is performed, namely the Peto & Peto modification of the Gehan-Wilcoxon test. This test place a higher weight on earlier survival differences. 
```{r gehan-wilcoxon test}
survdiff(Surv(time = time, event = status) ~ trt, # comparing both treatment groups
         data = veteran,
         rho = 01) # rho = 1 for different test; scalar parameter
```

This test produces a different result although it is still not significant, $\chi^2(1) = 0.9, p = .4$.


### Stratified tests

In the nonparametric applorach (Kaplan-Meier estimate) it is also possible to controll for a covariate. As long as the covariate is categorical with a small number of levels, regression-based approaches are not by all means necessary. The stratified log-rank test can be used instead.
```{r stratified test}
survdiff(Surv(time, status) ~ trt + 
           strata(celltype),  # stratified categorical variable celltype (survival::strata())
         data = veteran)
```




