---
title: 
author: "Felix Eßer"
date: "`r Sys.Date()`"
output: 
    html_document:
      toc: true
      toc_float: true
      number_sections: true
bibliography: '`r file.path(gsub("(?<=Statistical_Analyses).*", "", getwd(), perl = TRUE), "bibliography.bib")`'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Multiple Testing refers to the situation in which one conducts several statistical inferences.
It is of interest that the null hypothesis can be safely rejected which means that no $\alpha$-error is made. Therefore, the $\alpha$-error rate is conventionally set to $\alpha = .05$. However, when conducting INDEPENDENT multiple tests


A priori (also planned comparison) vs. a posteriori (also post-hoc)




Terminology:

- alpha
    - false discovery
- family-wise error rate (FWER)
    + also called experimentwise error rate
- false discovery rate (FDR)
- Multiple Testing procedures (adjusting p-values)
    - One-Step
        - all $p$ values are compared to a predetermined cut-off
    - Step-down
        - p values are evaluated from smallest to largest
        - each p value is compared to its own cut-off
        - if a p-value is not significant, the null-hypothesis is accepted for the corresponding test as well as all remaining tests having higher p-values (even if higher p value would be sig. )
    - Step-Up




family-wise error rate ($FWER$; also called experiment-wise error rate) is the probability of making at least one $\alpha$-error (type I error) in a set of statistical tests:
$$ FWER = \alpha_{fam} = 1 - (1-\alpha_r)^m$$
Here, $\alpha_r$ is the the $\alpha$-level for a specific test $r$ and $m$ denotes the number of all tests (see appendix for derivation).

```{r packages, message=FALSE}
library(tidyverse)
```

```{r fwer}
# vector m representing total number of tests
m <- 1:50
# alpha represents alpha for one test
alpha <- 0.05
# family-wise error-rate (fwer) dependent on m and alpha
fwer <- 1 - (1 - alpha)^m


ggplot(mapping = aes(x = m, y = fwer)) +
  geom_point(size = 1) +
  scale_x_continuous(breaks = seq(0, 50, 5)) +
  scale_y_continuous(breaks = seq(0, 0.90, 0.20)) +
  labs(title = bquote(
    atop("Family-Wise Error Rate ("*italic("FWER")*") Dependent on the Number of Total Tests ("*italic("m")*")",
         "with"~alpha == .(alpha))),
    x = expression(paste("Number of total tests (", italic("m"), ")")),
    y = expression(paste("Family-Wise Error Rate (", italic("FWER"), ")"))) +
  theme_minimal()


```

For $m = 20$ tests, the probability to have at least one false-significant result (with a significance level $\alpha = .05$ for each test) is $ FWER = \alpha_{fam} = 1 - (1-.05)^{20} = 0.64$. This means there is a 64% chance of getting at least one false-significant result. With $m = 20$ test, one would expect to have $m \cdot \alpha = 20 \cdot 0.05 = 1$ falsely rejected null-hypothesis (also called false discovery). As the figure shows, $FWER$ increases with the number of tests ($m$).


- What and why
    + Also knwon as 
        - multiple comparison (but closer to ANOVA and mean differences)
    + Multiple Testing refer to the situation where one is interested in 
    + $\alpha$ error inflation (link to alpha error)
        + what is a p value (short definition and link)
        + $p$-value is defined as the probability of obtaining a test statistic at least as large as the one observed, if the null hypothesis is true
    + family-wise error-ate and false discovery rate
- Where/Contexts
    + ANOVA 
    + correlations
    + but applicable anywhere
- How 
    + p-Value Adjustment
        - more independent from statistical analysis
        - applicable everywhere, where $p$ values are
    + Resampling Methods

    
    
# $p$-Value Adjustment

## Bonferronig correction

- divide the $\alpha$-level by the $m$ tests to get the new $\alpha$ level for all $r$ tests

Advantages:
- easy 
Disadvantages:
- very conservative
    + for many tests 

## Adjustment of critical values in mean comparisons 
- only relevant for mean comparisons (ANOVA)

- maybe decide to include it elsewhere (Comparing Means)
    + at least make a link
    
    
Tukey HSD and Tukey-Kramer
  - TukeyHSD only for equal sample sizes
  - Tukey Kramer for unequal sample sizes

```{r}

library(emmeans)
dat <- datasets::ToothGrowth |> 
  mutate(dose = factor(dose))

aov_mod <- aov(len ~ dose, data = dat)
summary(aov_mod)

TukeyHSD(aov_mod)

pairs(emmeans(aov_mod, ~ dose), adjust = "tukey") |> str()
# --> no CI

library(multcomp)

summary(glht(aov_mod, linfct = mcp(dose = "Tukey"), alternative = "greater"))

```




# Appendix

The formula for the family-wise error rate ($FWER$) can be derived from the multiplication theorem of probabilites (link to probability):
For two independent evens, event $A$ and event $B$, the probability that both event 
$$P(A \cap B) = P(A) \cdot P(B)$$

# References
