---
title: "Comparing Means - t-Test"
author: "Felix Eßer"
date: "10/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction


Comparing two means.

Effect sizes:
- cohen's $d$ is positively biased, hedges' $g$ is a bias corrected version for small sample sizes.




# Packages

```{r packages}
library(effectsize) # for effect size
library(skimr) # descriptive statistics
library(tidyverse) # data wrangling, plotting, etc.
```


# Comparing an observed mean to an expected mean - one sample $t$-test

# Comparing two independent groups - independent two sample $t$-tests
# Comparing two dependent groups - dependent two sample $t$-test

## Assumptions of the dependent $t$-test

## Conducting the dependent $t$-test

## Reporting the dependent $t$-test

## Visualizing Results

# Appendix: Theory and Mathematics of the $t$-tests


# Comparing an observed mean to an expected mean

# Comparing two independent groups

# Comparing two dependent groups






















