---
title: "Comparing Means - t-Test"
author: "Felix Eßer"
date: "10/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction


Comparing two means.

Effect sizes:
- cohen's $d$ is positively biased, hedges' $g$ is a bias corrected version for small sample sizes.




# Packages

```{r packages}
library(effectsize) # for effect size
library(skimr) # descriptive statistics
library(tidyverse) # data wrangling, plotting, etc.
```


# Comparing an observed mean to an expected mean - one sample $t$-test

# Comparing two independent groups - independent two sample $t$-tests
# Comparing two dependent groups - dependent two sample $t$-test

The means on a metric outcome of two dependent samples can be compared with the $t$-test for dependent samples. Dependent samples can arise through repeated measurement of the same cases. As long as one measurement of one sample can be connected to a measurement of the other sample, dependent samples are present. Therefore, spouses are also an example of dependent samples.

If the comparison of more than two dependent samples are of interest, the ANOVA for dependent samples can be used.

At first the dataset will be created.

```{r dependent-data}
# Load dat
dep_data <- data.frame(
  pre = c(9, 8, 1, 3, 2),
  post = c(16, 11, 15, 12, 9)
) %>% as_tibble()

# Show data
dep_data
```

Now, we get the descriptive statistics for the two dependent samples, as well as the difference. These will be incorporated into the report of the $t$-test for dependent samples (below). The mean and standard deviation are outputted along with other statistics by aid of the `skim()` function of the `{skimr}` package.

```{r descriptive-dep}
# Calculate descriptive statistics 
dep_descr <- 
  dep_data %>% 
  mutate(difference = post - pre) %>% # create difference variable
  skim()

# Output descriptive statistics
dep_descr
```



## Assumptions of the dependent $t$-test

## Conducting the dependent $t$-test

## Reporting the dependent $t$-test

## Visualizing Results

# Appendix: Theory and Mathematics of the $t$-tests


# Comparing an observed mean to an expected mean

# Comparing two independent groups

# Comparing two dependent groups






















