---
title: "Multivariate Analysis of Variance (MANOVA)"
author: "Felix Eßer"
date: "1/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The multivariate analysis of variance (MANOVA) is a statistical procedure to compare groups in regard to means on more than one outcome variable. 

Why MANOVA:

- $\alpha$ error cumulation through multiple testing
  + in conjunction: greater power (however besides intercorrelation of the outcomes also the effect sizes influence power)
- respects relationships among the outcome variables
  + possible detection whether groups differ along a combination of dimensions

Reasons to conduct multiple ANOVAs:

- outcome variables cannot be put in a joint theoretical framework (no relationship among the outcome variables)
- obtain exploratory group differences
- replication of former ANOVA results


## Theory of MANOVA

### Decomposition of measured scores

SSCP = sum of squares and cross product matrix

$\mathbf{T}$ = Total SSCP  
$\mathbf{B}$ = Between SSCP  
$\mathbf{W}$ = Within SSCP  

Indices:  
$m$ = Index of person, $m = 1,..., N$, where $N$ is the number of persons  
$i$ = Index of outcome variable, $i = 1,..., p$, where $p$ is the number of outcome variables  
$j$ = Index of group, $j = 1,..., k$, where $k$ is the number of groups  

Therefore, $x_{mij}$ denotes the value from person $m$ on outcome variable $i$ in group $j$. 

$$\mathbf{T} = \mathbf{B} + \mathbf{W}$$
$$ \begin{pmatrix}
  SS_{T~1,1} & CP_{T~1,2} & \cdots & CP_{T~1,p} \\
	CP_{T~2,1} & SS_{T~2,2} & \cdots & CP_{T~2,p} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	CP_{T~p,1} & CP_{T~p,2} & \cdots & SS_{T~p,p} 
  \end{pmatrix} = 
  \begin{pmatrix}
  SS_{B~1,1} & CP_{B~1,2} & \cdots & CP_{B~1,p} \\
	CP_{B~2,1} & SS_{B~2,2} & \cdots & CP_{B~2,p} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	CP_{B~p,1} & CP_{B~p,2} & \cdots & SS_{B~p,p} 
  \end{pmatrix} +
    \begin{pmatrix}
  SS_{W~1,1} & CP_{W~1,2} & \cdots & CP_{W~1,p} \\
	CP_{W~2,1} & SS_{W~2,2} & \cdots & CP_{W~2,p} \\
	\vdots  & \vdots  & \ddots & \vdots  \\
	CP_{W~p,1} & CP_{W~p,2} & \cdots & SS_{W~p,p} 
  \end{pmatrix}$$



Calculation of the elements of the total SSCP:  
$\displaystyle SS_{T~i,i} = \sum_{j = 1}^{J}\sum_{m=1}^{n_j} \left( x_{mij} - \overline{x}_{i} \right)^2$  
$\displaystyle CP_{T~h,i} = \sum_{j=1}^J\sum_{m=1}^{n_j} \left(x_{mhj} - \overline{x}_h\right) \cdot \left(x_{mij} - \overline{x}_i \right)$,    where $h \in [1,p] \land h \neq i$ 

The $SS_{T~i,i}$ denotes the total sum of squares. It is a unstandardized measure for the interindividual variability on one outcome variable irrespective of the groups.  
The $CP_{T~h,i}$ denotes the total sum of cross products. It is a unstandardized measure capturing the relationship between two outcome variables irrespective of the groups. 

Calculation of the elements of the between SSCP:
$\displaystyle SS_{B~i,i} = \sum_{j = 1}^{J}{n_j~\cdot~\left(\overline{x}_{ij}  - \overline{x}_i\right)^2}$  
$\displaystyle CP_{B~h,i} = \sum_{j=1}^J n_j \cdot \left(x_{hj} - \overline{x}_h\right) \cdot \left(x_{ij} - \overline{x}_i \right)$,    where $h \in [1,p] \land h \neq i$ 

The $SS_{B~i,i}$ denotes the between sum of squares. It is a unstandardized measure for the variability of one outcome variable between the groups.  
The $CP_{B~h,i}$ denotes the between sum of cross products. It is a unstandardized measure capturing the relationship between two outcomes group means deviation from the grand mean.

Calculation of the elements of the within SSCP: 
$\displaystyle SS_{W~i,i} = \sum_{j = 1}^{J}\sum_{m=1}^{n_j} \left( x_{mij} - \overline{x}_{ij} \right)^2$  
$\displaystyle CP_{W~h,i} = \sum_{j=1}^J\sum_{m=1}^{n_j} \left(x_{mhj} - \overline{x}_{hj}\right) \cdot \left(x_{mij} - \overline{x}_{ij} \right)$,    where $h \in [1,p] \land h \neq i$ 

The $SS_{W~i,i}$ denotes the within sum of squares. It is a unstandardized measure for the variability of one outcome variable within the groups.  
The $CP_{B~h,i}$ denotes the with sum of cross products. It is a unstandardized measure capturing the relationship between two outcomes within groups.


### Hypotheses and test statistics



### Discriminant function


## Load packages and data

```{r packages, results='hide', message=FALSE}
library(biotools) # Box-M test
library(skimr)
library(tidyverse)
```

```{r data}
# Import data from psych package on ability measures
sat_data <- psych::sat.act %>% 
  as_tibble() %>% 
  mutate(gender = factor(gender, labels = c("male", "female")),
         education = factor(education))
  
glimpse(sat_data)
```
| Variable  | Meaning                                                        |
|-----------|----------------------------------------------------------------|
| gender    | males = 1, females = 2                                         |
| education | self reported education 1 = high school... 5 = graduate work   |
| age       | in years                                                       |
| ACT       | ACT composite scores may range from 1 -- 36; Norms: mean of 20 |
| SATV      | SAT verbal scores may range from 200 -- 800                    |
| SATQ      | SAT quantitative scores may range from 200 -- 800              |


 The following model will be analyzed: `ACT`, `SATV` and `SATQ` are the outcome variables and `education` is the grouping variable.


# Assumptions of MANOVA
1. Independent residuals
    - within and between groups
2. Multivariate normality of the residuals
    - can be tested with Merida's test of multivariate normality
3. Homogeneity of the residual covariance matrices
    - this includes homoskedasticity (homogeneity of variance = equal residual variances) and homogeneity of the residual covariances between the groups
    - can be tested with the Box-M Test

## Testing the Assumptions

### 1. Independent residuals
### 2. Multivariate normality of the residuals
### 3. Homogeneity of the residual covariance matrix

```{r box-m-test}
#boxM(sat_data[, c("ACT", "SATV", "SATQ")], factor(sat_data$gender)) # does not work
```


# MANOVA

```{r}
manova_mod <- manova(cbind(ACT, SATV, SATQ) ~ education, data = sat_data)
sapply(c(1,2,3), function(x) sum(1,x))


sapply(c("Pillai", "Wilks", "Hotelling-Lawley", "Roy"), function(x) summary(manova_mod, test = x))

walk(c("Pillai", "Wilks", "Hotelling-Lawley", "Roy"), ~summary(manova_mod, test = .x))

tests <- c("Pillai", "Wilks", "Hotelling-Lawley", "Roy")
for (name in 1:length(tests)) {
  print(name)
  print(summary(manova_mod, test = tests[[name]]))
}

summary(manova_mod, test = "Pillai")

```

# Discriminant Function Analysis


# Combined Report









