---
title: "Nonparametric Comparison of Means"
author: "Felix Eßer"
date: "3/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

```{r packages}
library(coin)
library(tidyverse)
```


# Comparison of two independent groups - Fisher-Pitman-Test

```{r}
ind_data <- tibble(value = c(18, 24, 25,
                             21, 29, 29, 30, 31, 31, 31),
                   grp = factor(rep(c(1, 2), c(3,7))))
ind_data

oneway_test(value ~ grp, data = ind_data, alternative = "less")
oneway_test(value ~ grp, data = ind_data, alternative = "less", distribution = "asymptotic")
oneway_test(value ~ grp, data = ind_data, alternative = "less", distribution = "exact")

```



# Comparison of two dependent groups - Fishers randomization test for 2 dependent variables


Comparison of Fisher Test with:

- Sign-Test (create link)
    + sign test only considers frequency of negative and positive differences
    + sign test does not take into account size of differences
- Wilcoxon signed-rank test (create link)
    + weighs signs of differences with corresponding ranks
- Fishers randomization test
    + wighs signs of differences with numerical differences 
    
--> Fishers randomization test uses the most of the data but it has to be applicable.


## Assumptions
1. Independent observation of paired observation 


```{r}
dep_data <- tibble(id = factor(1:10), 
       pre = c(17, 22 , 22, 15, 24, 22, 21, 21, 17, 21),
       post = c(20, 21, 21, 22, 24, 22, 23, 21, 22, 23)) %>% 
  pivot_longer(c(pre, post), names_to = "time", 
               names_transform = list(time = as.factor))

dep_data
oneway_test(value ~ time | id, data = dep_data, distribution = "exact", alternative = "greater")
```


# Comparision of several independent groups - Oneway randomization ANOVA


