---
title: "Exploratory Data Analysis"
author: "Felix Eßer"
date: "10/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis and Data Screening

In advance to statistical (inferential) analyses, it is important to conduct exploratory data analysis (EDA) or data screening. Through EDA it is possible to get a *feeling* for the data, detect outliers, assess assumptions, etc. Additionally, analyses are performed in the `base R` framework and the `tidyverse` framework. Of course there are several packages to conduct these kind of analyses. In the following, the packages `skimr` and `psych` are used. The package `skimr` can be used in conjuction with the `tidyverse` framework and is kableable. In contrast, the package `psych` provides additional statistics. There are a bunch of packages which can be used to conduct basic descriptive statistics like `pastecs`, `Hmisc`, `summarytools`, and surely others. 

### Packages and datasets

```{r load packages}
library(tidyverse) # data manipulation and plots
library(skimr)
```

Throughout the dataset `mtcars` is used to demonstrate the proposed concepts. 

```{r data}
data("mtcars") # from tidyverse package
mtcars <- mtcars %>% as_tibble() %>% 
  # manipulate tibble to obtain different classes of columns
  mutate(vs = factor(vs, labels = c("v-shaped", "straight")),
         character = rep(c("A", "B"), length.out = nrow(mtcars)))
```



## Univariate Exploratory Data Analysis
This is used to check the accuracy of the input. 

## Quick Exploratory Data Analysis on the whole dataset

### `Base R` approach
In `base R` the `summary()` function can be used:
```{r base R}
summary(mtcars)
```

### `skimR` approach

The function `skim` can be used in pipes and can itself be the input of a pipe in which case the summary data can be further manipulated.

```{r skim}
mtcars %>% 
  group_by(am) %>% # skim recognises grouped tibbles
  skim()
```

With the function `skim_tee()` it is possible to obtain an output from a pipe but use the data in a further pipeline. 
Using the function `skim_with()` additional/other statistics can be calculated. However, this procedure is a bit unconvenient as `skim_with()` is a closure and returns therefore a function which has to be assigned.
```{r skim_with}
skim_median <- skim_with(numeric = sfl(median), # sfl = skimmer_function_list
                         append = FALSE) # no default statistics
skim_median(mtcars)
```

### `psych` approach

Using the function `describe` from the package `psych` results more easily in additional statistics.
```{r describe}
mtcars %>% 
  select(mpg, vs, character) %>% 
  describe() 
```

The output of the `describe()` function can be used in further pipes. The advantage is that statistics regarding the distribution are given (skew and kurtosis).
To obtain descriptive statistics by group, the function `describeBy()` can be used. As can be seen, the output of the `describeBy` function is not well knitable.
```{r describeBy}
mtcars %>% 
  select(mpg, vs, am) %>% 
  describeBy(group = .$am) # short: describeBy(mtcars ~ am)
```



## Measures of Central Tendency
Measures of central tendency are the mode, median and mean. In addition, the weighted mean, the geometric mean and the harmonic mean are 

## Measures of dispersion


## Assumption of normality


## Bi- and Multivariate Exploratory Data Analysis

```{r marida}
marida()
```



## Missing data

