---
title: "Exploratory Data Analysis"
author: "Felix Eßer"
date: "10/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis and Data Screening

In advance to statistical (inferential) analyses, it is important to conduct exploratory data analysis (EDA) or data screening. Through EDA it is possible to get a *feeling* for the data, detect outliers, assess assumptions, etc. Additionally, analyses are performed in the `base R` framework and the `tidyverse` framework. Of course there are several packages to conduct these kind of analyses. In the following, the packages `skimr` and `psych` are used. The package `skimr` can be used in conjuction with the `tidyverse` framework. There are a bunch of packages which can be used to conduct basic descriptive statistics like `pastecs`, `Hmisc`, `summarytools`, and surely others. 

```{r load packages}
library(tidyverse) # data manipulation and plots
library(skimr)
```


## Univariate Exploratory Data Analysis

This is used to check the accuracy of the input. 



## Assumption of normality


## Bi- and Multivariate Exploratory Data Analysis

## Missing data

