---
title: "Tests for Categorical Data Analysis"
author: "Felix Eßer"
date: "10/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

```{r load packages, message=FALSE}
library(rstatix)
library(tidyverse)
library(vcdExtra)
```


## Comparing an observed frequency distribution with an exepected frequency distribution

### dichotomous data
binomial test, chi-square test
### polytomous data
multinomial test, chi-square test

## Comparing independent groups regarding frequencies

### Comparing two independent groups regarding frequencies
#### dichotomous data
fisher-yates-test, chi-square test
#### polytomous data
freeman-halton-test, chi-square-test

### Comparing several independent groups regarding frequencies
chi-square-test, fuchs-kennet-outliertest

## Comparing dependent groups regarding frequencies
### Comparing two dependent groups regarding frequencies
#### dichotomous data
McNemar-test, 
#### polytomous data
Marginalhomogenitätstest nach Lehmacher, Bowker-symmetry-test

### Comparing several dependent groups regarding frequencies
#### dichotomous data
Cochran's Q test
#### polytomous data











