---
title: "Power Analysis"
author: "Felix Eßer"
date: "11/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Disclaimer: All informations are extracted from the vignette of the `pwr` package. 

Four quantities are interrelated:
1. sample size $n$
2. effect size $\delta$
3. significance level $\alpha$
4. power $1-\beta$


Different types of power analysis:

* A priori power analysis
    + determine required sample size $n$
    + determine $1-\beta$ (sometimes referred to as post-hoc power analysis)
* Post-hoc power analysis
    + effect size is given from empirical data of own study ($\hat{\delta}$; post-hoc or a posteriori)
    + assess probability of revealing given effect size on the assumption it really exists


## Effect Sizes
Different effect sizes exist and the correct effect size has to be used to obtain correct results. 


The following table has to be adjusted: 
| Test                      | small | medium | large |
|---------------------------|-------|--------|-------|
| tests for proportions (p) |   0.2 |    0.5 |   0.8 |
| tests for means (t)       |   0.2 |    0.5 |   0.8 |
| chi-square tests (chisq)  |   0.1 |    0.3 |   0.5 |
| correlation test (r)      |   0.1 |    0.3 |   0.5 |
| anova (anov)              |   0.1 |   0.25 |   0.4 |
| general linear model (f2) |  0.02 |   0.15 |  0.35 |


# The `pwr` package
    
```{r packages}
library(pwr)
```
 
    
The package `pwr` contains different functions 

| `R` Function     | Power Analysis for Statistical Test                |
|------------------|----------------------------------------------------|
| `pwr.p.test`     | one-sample proportion test                         |
| `pwr.2p.test`    | two-sample proportion test                         |
| `pwr.2p2n.test`  | two-sample proportion test (unequal sample sizes)  |
| `pwr.t.test`     | two-sample, one-sample and paired t-tests          |
| `pwr.t2n.test`   | two-sample t-tests (unequal sample sizes)          |
| `pwr.anova.test` | one-way balanced ANOVA                             |
| `pwr.r.test`     | correlation test                                   |
| `pwr.chisq.test` | chi-squared test (goodness of fit and association) |
| `pwr.f2.test`    | test for the general linear model                  |


The functional idea for the functions of the `pwr` package is to leave out the argument (of the four interrelated quantities) one wants to calculate.



# Using the `pwr` package

## Power Analysis for the general linear model

The power analysis for the general linear model refers to the overall $F$-test, testing the null hypothesis $H_0: R^2 = 0$. Therefore, besides the standard quantities, the following quantities have to be used in the power analysis:
* $df_{numerator}$ of the F-test
    + equals the number of coefficients minus the intercept
* effect size $f^2$ 

For the analysis the caluclation of the $df_{denominator}$ is crucial: 

$df_{denominator} = N - df_{numerator} - 1$.

Therefore, the equation can be solved to $N$, which is: 

$N = df_{numerator} + df_{denominator} + 1$.

```{r glm}
power_glm <- pwr.f2.test(u = 2, # df numerator (here 3 predictors with intercept)
                         f2 = 0.3/(1 - 0.3), # R^2 = 0.3
                         sig.level = 0.05, # alpha = 0.05
                         power = 0.8)
power_glm
```
Using the function to calculate $N$, $N$ equals $2 + 23 + 1 = 26$.



# Report

Example report from "Statistical Methods for Psychology" (Howell)

"Based on the work of Jones and others (list references) we estimated that our mean difference would be approximately 8 points, with a standard deviation within each of the groups of approximately 5. This would give us an estimated effect size (d) of 8/11 5 .73. We were aiming for a power estimate of .80, and to reach that level of power with our estimated effect size, we used 30 participants in each of the two groups."


