---
title: "Correlation"
author: "Felix Eßer"
date: "`r Sys.Date()`"
output: 
    html_document:
      toc: true
      toc_float: true
      number_sections: true
bibliography: '`r file.path(gsub("(?<=Statistical_Analyses).*", "", getwd(), perl = TRUE), "bibliography.bib")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction and Overview

Correlation coefficients quantify the degreee of covariation between two variables. Due to the possible combinations of different scales of measurement, several correlation coefficients exist. The presentation of the different correlation coefficients is structured along the different combinations of scales of measurement. 

For each correlation coefficient the general outline is as follows:
1. Assumptions of the correlation coefficient
2. Computation
3. Interpretation
4. Properties
5. Visualization

In order to have clear overview over the several correlation coefficients, the mathematical background is located in the appendix. 


# Load Packages and Data

```{r packages}
library(tidyverse) # data wrangling, etc. 
```



```{r data}
glimpse(mtcars)
```



# Two Nominal Variables

## Two Dichotomous Variables

### The $\varphi$ coefficient - Product-Moment-Correlation of dichotomous variables

value range: [-1, +1]
boundary value can only be obtained if relationship is perfect **and** both variables have the same distribution.

The more both distributions differ, the smaller the possible absolute value of $\varphi$.

There are corrections but it is easier to use other correlation measures for dichotomous variables.

### Yules Q

### Odds-Ratio

### Tetrachoric Correlation

Application of the polychoric correlation on dichotomous variables.


## Two Polytomous Variables

### Contingency Coefficient

Based on $\chi^2$ value. 
inferior to Cramer's $V$ becauseit can never reach it's upper boundary!

### Cramer's $V$

Based on the $\chi^2$ value
is identical to $\varphi$ in the case of dichotomous variables. 
value range: [0, 1]



## A Dichotomous and a Polytomous Variable

### Cramer's $V$


# Two Ordinal Variables

Kendall's $\tau$
-> used for singular ordinal variables without tied ranks

$e$ coefficient
--> used for singular ordinal variables with tied ranks


$\gamma$ coefficient
-> used for ordinal variables with ordered categories


polychoric correlation
-> used for ordinal variables with ordered categories
-> assumes that metric normally distributed variables underlie both ordinal variables with 


Somer's $d$
-> used for a singular ordinal variable (can have ties) and a ordinal variable with ordered categories

# Two Metric Variables

## Pearson's Product-Moment-Correlation

value range: $[-1, +1]$



# A Nominal and a Ordinal Variable

# A Nominal and a Metric Variable

dichotomous nominal variable:
point-biserial correlation
biserial correlation

polytomous nominal variable:
eta = square root of eta squared (from ANOVA)
-> eta for dichotomous nominal variable corresponds with point-biserial correlation


# A Ordinal and a Metric Variable

ordinal variable with ordered categories
polyserial correlation
- assumes that a metric variable underlies the ordinal variable


singular ordinal variable








# Appendix: Theory and Mathematics of the Correlation Coefficients




# Two Metric Variables

$$CP_{XY} = \sum^n_{i = 1} (x_i - \bar{x}) \cdot (y_i - \bar{y})$$



