---
title: "Partial_Correlation"
author: "Felix Eßer"
date: "10/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Correlation between more than two variables

Reasons for usage
- Reveal redundant relationships
  - How much variance in $Y$ is explained by $X_1$ independent from $X_2$?
- Reveal spurious correlations
  - Can the relationship between two variables be expressed by a third variable?
- Reveal masked relationships
  - Is a relationship between two variables is masked (covered up) by a third variable


Usually shown for product-moment correlation but also possible for other variable types: Add this!
- categorical variables: log-linear analysis
- ordinal variables: similar to shown below possible with Kendall's $\tau$



Usage
- partial correlation
- semi-partial correlation



## Partial Correlation

- is a bivariate correlation between two residual variables



zero order correlation
- bivariate correlation (between two variables)
- no additional variable is partialized out


first order (partial) correlation
- bivariate correlation in which a third variable was partialized oout


Concept
- partial correlation is the correlation of residual variables
- $r_{YX\bullet Z} = r_{E_{X|Z}E_{Y|Z}}$
- variable that was partialized out comes after the bullet



conditional correlation
- partial correlation can be interpreted as an average conditional correlation: $r_{XY|Z=z}$
- averaged over the distribution of variable $Z$
- averaged by reference to the conditonal variances and covariances






$$r_{YX \bullet Z} = \frac{r_{YX} - r_{YZ}\cdot r_{XZ}}{\sqrt{\left(1-r_{YZ}^2\right) \left(1-r_{XZ}^2\right)}}$$


- $r_{YX\bullet Z}$ = partial correlation of 
- $r_{YX}$ = zero order correlation of $X$ and $Y$
- $r_{YZ}$ = zero order correlation of $Z$ and $Y$
- $r_{XZ}$ = zero order correlation of $Z$ and $X$


Add path diagram illustrating partial correlation




Properties of the partial correlation

- value falls between -1 and +1 (equivalent to the bivariate correlation)


Interpretation

1) $r_{XY} = r_{XY\bullet Z}$
  - zero-order correlation is equal to the first-order partial correlation
  - if $Z$ is uncorrelated with $X$ and $Y$
2) $|r_{XY}| < |r_{XY\bullet Z}|$
  - absolute value of the zero-order correlation is smaler than the absolute value of the first-order partial correlation
  - $Z$ variable can explain (part of) the relationship (phrase this better)
  - special case: $r_{XY\bullet Z} = 0$
    - product of correlation of variables $X$ and $Y$ with variable $Z$ ($r_{YZ}\cdot r_{XZ}$) is equal to the zero-order correlation ($r_{YX}$; see formula above)
3) $|r_{XY}| > |r_{XY\bullet Z}|$
  - if $Z$ is uncorrelated with either $X$ or $Y$ but not both



Add examples with numbers
- for each example provide a zero-order correlation matrix 
- than provide partial correlations





#### Suppressor Variable ####

- must be restructred so it aligns with the enumeration above

3) $|r_{XY}| > |r_{XY\bullet Z}|$
  - if $Z$ is uncorrelated with either $X$ or $Y$ but not both


possible that partial correlation has opposite sign than zero-order correlation


Add sick tikz venn diagrams
- unfortunately, it seems that it is not possible with R because shading areas with lines seems to be difficult
- Add multiple venn diagrams in one figure representing different concepts
- zero-order correlations between X and Y as well as Z and Y
- squared partial correlation: $r_{XY\bullet Z}^2$
- squared semi-partial correlation $r_{X
- Explanation venn diagram
  - each circle represents the variance of a variable
  - overlap represents the shared variance between two variables: shared variance = squared correlation

```{r}
library(VennDiagram)
grid.newpage() # create new plotting page
draw.triple.venn(
    area1 = 10, area2 = 10, area3 = 10,
    n12 = 2, n13 = 2, n23 = 2, n123 = 2,
    category = c("X", "Y", "Z"),
    fill = c("blue", "red", NA),
)

```




### Squared Partial Correlations ###

- square of a partial correlation is a proportional reduction in error
- represents the proportion of the variation in $Y$ that is explained by $X$, out of that left unexplained by $Z$ (Change this definition)


$$r_{YX\bullet Z}^2 = \frac{R_{Y|X,Z}^2 - r_{YZ}^2}{1-r_{YZ}^2}$$

- $R_{Y|X,Z}^2$ = determination coefficient for a multiple regression in which $Y$ is regressed on $X$ and $Z$
- $r_{YZ}^2$ = squared zero order correlation between $Z$ and $Y$
- $R_{Y|X,Z}^2 - r_{YZ}^2$ = partial proportion of variance in $Y$ uniquely explained by $X$
- $1- r_{YZ}$ = proportion of variance in $Y$ unexplained by $Z$


### Higher-Order Partial Correlations ###

second-order partial correlation

### Assumptions for Partial Correlations ###





### With R


Testing whether zero-order correlation differs significantly from partial correlation


## Semi-Partial Correlation

- third variable is only partialized out of one variable

correlation of variable X with residual variable Y for which variable Z was partialized out


$$r_{(Y\bullet Z)X} = \frac{r_{XY} - r_{XZ} \cdot r_{YZ}}{\sqrt{1-r_{YZ}^2}}$$
